Computational design of RNA-based oscillatory circuits

### 2015-09-07 MScWriteup.tex
making illiana's changes, beginning to make manish's

\begin{figure*}[t]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a 'double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}

@@

In this system, transcription of the designed sRNA and mRNA are placed under the control of promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$ \cite{Lutz1997}. These are in turn controlled by two transcriptional repressors, TetR and LacI, which are naturally present in the strain of \text{E. coli} considered. These repressors disable the promoter regions, and so by default transcription of both RNA's is turned off, and no protein is produced. 
These repressors can themselves be disabled by the presence of two chemicals, aTc and IPTG, which can be introduced externally into the cell (Fig. \ref{ANDGate}).  So transcription of the two RNA's is indirectly controlled by the presence of two chemicals - if neither is present, sRNA and mRNA transcription is repressed, and no protein is produced. If only one is present, the AND gate remains off, either because there is no mRNA to be translated into protein, or because the mRNA is self repressed. But when both are present, the conformational change discussed above occurs, and protein is produced.

@@

mRNA concentrations in this system shown in Fig. \ref{ANDGate} can be indirectly observed by designing the mRNA to code for GFP \footnote{ The methodology of \cite{Rodrigo2012} only optimises the 5' UTR of the mRNA, so the actual protein being coded for is unimportant}. Recent experiments have used timelapse microscopy to observe the fluoresence of bacteria which contain the above sRNA-mRNA pair, as they are periodically forced with a varying aTc or IPTG concentration \cite{Jaramillo}.

The bacteria are grown a single layer thick in rows of chambers. A medium constantly flows through these chambers, allowing normal feeding of the bacteria, and the introduction of aTc or IPTG. The chambers are monitored with software which traces the position of each cell over time, allowing timeseries of individual cell fluorescences to be recorded.

\begin{figure}[h]
\centering
\includegraphics[trim = 0 0 0  85 ,page=17 ,clip = true, scale=0.3]{WCPM_Warwick-talk45-220115.pdf}
\centering
\includegraphics[trim = 0 0 0  90 ,page=18 ,clip = true, scale=0.3]{WCPM_Warwick-talk45-220115.pdf}
\caption{A schematic of an experimental setup which allows single cell fluorescences to be recorded over time. Bacteria are grown in chambers, with a constant supply of medium flowing through them, which allows the introduction of aTc and IPTG. Software then traces these cells, allowing fluorescence timeseries to be built.  Image reproduced from \cite{Jaramillo}.}
\label{}
\end{figure}

The data we will consider consists of two sets of individual cell timeseries, labelled \textit{13\_9}, \textit{14\_7}. They correspond to different experimental runs of the above apparatus, for which IPTG concentration was held constant, at a value assumed large enough to saturate the cell's response, and aTc concentrations were varied periodically. Section \ref{Initial  Experimental Data} shows the full datasets, with their forcing functions.

@@

We propose a simple model, consisting of a set of ODE's with mass action kinetics \cite{UriAlon}, to describe the system. Tables \ref{ModelParameters}, \ref{StateVariables} give complete descriptions of the parameters and state variables.

\begin{align}
\frac{ds}{dt} &= \frac{N\alpha_{T}}{f_{T}} y(t)-(\mu + \delta_{s})s -k_{\mathrm{on}}sm +k_{\mathrm{off}}s:m \label{eq:s}\\
\frac{dm}{dt} &=  \frac{N\alpha_{L}}{f_{L}}x(t)-(\mu + \delta_{m})m -k_{\mathrm{on}}sm +k_{\mathrm{off}}s:m  \label{eq:m}\\
\frac{ds:m}{dt} & = k_{\mathrm{on}}sm  - (k_{\mathrm{off}}+ k_{\mathrm{hyb}})s:m  -(\mu + \delta_{sm} )s:m \label{eq:sm}\\
\frac{dc}{dt} & = k_{\mathrm{hyb}}s:m  -(\mu + \delta_{c})c  \label{eq:c} \\
\frac{dp}{dt} & = \beta m +f_{s}\beta c -(\gamma + \mu + \delta_{g})p - \frac{v_{z}p}{K_{z}+p+g}  \label{eq:p} \\
\frac{dg}{dt} & = \gamma p - (\mu + \delta_{g})g - \frac{v_{z}g}{K_{z}+p+g} \label{eq:g} \\
z &= z_{0} +\frac{g}{\Theta} \label{eq:z}
\end{align}
  Based on the reaction mechanism in Fig. \ref{reactionscheme}, the hybridization of the sRNA and mRNA first into an unstable complex, then a stable one, is modelled in eqs. \ref{eq:s}-\ref{eq:c}. The initial binding is modelled as a reversible reaction with forward and backward rates $k_{on}$ and $k_{off}$.

\begin{align*}
\mathrm{
sRNA \chemsign+ mRNA\chemrel[$k_{on}$][$k_{off}$]{<>} sRNA:mRNA\textsubscript{unstable}
}
\end{align*}
after which the stabilization is modelled as an irreversible reaction with rate $k_{hyb}$.

\begin{align*}
\mathrm{
sRNA:mRNA\textsubscript{unstable} \chemrel[$k_{hyb}$]{->} sRNA:mRNA\textsubscript{stable} 
}
\end{align*}

 In addition, these complexes are given degradation rates, $\delta_{s}$, $\delta_{m}$, $\delta_{sm}$, $\delta_{c}$, and dilution of chemical concentrations due to cell growth are modelled with a dilution rate $\mu$.

@@

\begin{table}[h]
\renewcommand{\arraystretch}{1.3}
\caption{Model Parameters (those to be estimated shown in bold)}
\label{ModelParameters}
\centering
\begin{tabular}{| l | l | p{0.6\linewidth} |}
\hline \textbf{Parameter} &  \textbf{Units} & \textbf{Definition}  \\
\hline \hline N & & Number of copies of promoter existing on plasmid DNA  \\
\hline $z_{0}$ &  AFU & Baseline experimental fluorescence  \\
\hline $\alpha_{L}$ & nM/min & Maximal transcription rate of $P_\mathrm{LlacO-1}$ promoter\\
\hline $\alpha_{T}$  &  nM/min  & Maximal transcription rate of $P_{\mathrm{LtetO-1}}$ promoter \\
\hline $f_{L}$ &  & Unitless ratio between repressed and unrepressed $P_\mathrm{LlacO-1}$ transcription rate   \\ 
\hline $f_{T}$ &  & Unitless ratio between repressed and unrepressed $P_{\mathrm{LtetO-1}}$ transcription rate  \\
\hline $\delta_{g}$  & /min  & GFP degradation rate  \\
\hline $\gamma$ &  /min & GFP maturation rate  \\
\hline $v_{z}$ & nM/min & degradation constant of clpx  \\
\hline $K_{z}$   &   nM/min & Dissociation constant of clpx  \\
\hline $\boldsymbol{\Theta}$  &   nM/AFU & Ratio between GFP concentration and observed fluoresence  \\
\hline $\boldsymbol{\mu}$ &  /min & Dilution rate  \\
\hline $\boldsymbol{\delta_{m}}$ &  /min & mRNA degredation rate  \\
\hline $\boldsymbol{\delta_{s}}$ &  /min & sRNA degredation rate  \\
\hline $\delta_{sm}$ &  /min & Unstable sRNA:mRNA degradation rate  \\
\hline $\delta_{c}$ &  /min & Stable sRNA:mRNA degradation rate  \\
\hline $\boldsymbol{k_{on}}$ &   /min & sRNA:mRNA binding rate \\
\hline $\boldsymbol{k_{off}}$ &  /min & sRNA:mRNA unbinding rate \\
\hline $\boldsymbol{k_{hyb}}$ &  /min & sRNA:mRNA hybridization rate \\
\hline $\boldsymbol{\beta}$ &   /min & Baseline translation rate of repressed mRNA \\
\hline $\boldsymbol{f_{s}}$ & & Ratio of repressed mRNA to unrepressed complex translation rate. \\
\hline
\end{tabular}
\end{table}

@@

Our next goal is to estimate the unknown parameters of this model, given the available fluorescence time series data, by fitting predicted time series from the model to the data. Typically, this is done by minimising the least squares error between model prediction and the experimental data \cite{Brewer2008,Algorithms2003, Hu2015}. Suppose we have some ODE model of our system

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t)
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. An error between the model prediction and an experimental time series is defined as
\begin{align}
J(\boldsymbol{\theta}) = \sum_{i =1}^{N} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2
\end{align}
where the experimental timeseries, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots N$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a timeseries from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\min_{\boldsymbol{\theta}} \sum_{i =1}^{N} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function is multimodal, and may be very rugged. \footnote{this is true even if ODE model is linear in its parameters, as is almost the case for us - though the ODE model is linear, the resulting solutions are in general not. A counterexample is the harmonic oscillator.} A local optimisation algorithm will often get stuck in local minima. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular recommends several Evolutionary Algorithms, of which we choose one, the CMA-ES \cite{Hansen2006,Hansen2011}.

In order to reduce the dimensionality of our search space, we can perform a literature search for existing values of some of our parameters, simplify our model to remove others, and place bounds on those that remain. Section \ref{Parameter literature review} contains a list of parameter values found in the literature, where available, and their reference, as well as initial bounds placed on parameters not found in the literature. To further reduce the search space, we assume that $\delta_m$, $\delta_{sm}$ and $\delta_c$ all take similar values, and set them equal. After this is done, we are left with a 9 dimensional search space, bounded by a hypercube (parameters to be estimated are shown in bold in table \ref{ModelParameters}).

\begin{figure*}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.35, clip = true, trim = 80 0 80 0]{13_9_hist.eps}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset. }
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.38,clip = true, trim = 60 12 80 0]{14_7_hist.eps}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
    \end{subfigure}

        \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.25, clip = true, trim = 100 0 100 0]{13_9_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{13\_9} dataset. }
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.25, clip = true, trim = 80 0 80 0]{14_7_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{14\_7} dataset.}
    \end{subfigure}

    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.37]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset. A}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.37]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
    \end{subfigure}
    \caption{Parameter estimates, inter-parameter correlation values, and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off, and a level which saturates the cell's response}
\label{InitialResults}
\end{figure*}

@@

Though the predictions are reasonable, they are substantially worse than the predictions for the data they were trained on. The first reason for this may be experiment to experiment variability of parameter values \cite{Hu2015}. The second may be that the model is overfitting.

@@

These problems may stem from structural inadequacies in the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions \cite{Jimenez-Hornero2008,Grewal1976 }. If this is the case, no amount of experimental data will allow us to estimate parameters, and we must consider reformulating the model. \footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product can ever be estimated, a problem no amount of data can fix.}

Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in general this is not true. \footnote{\cite{Beck} gives an example in which a parameter only effects model predictions after several hours, though others will affect it at all timescales - in this example, if we only took data for a few minutes, the parameter would be inestimable, but in principle it is not.} In this case, parameter estimates be may improved by taking data in more varied experimental conditions, and attempting to observe as many components of the model output as possible.

@@

It can be shown that if sensitivity co-efficients are linearly dependent over the range of observation values, the associated parameters cannot be simultaneously estimated \cite{Beck}. Related to this fact, a number of measures have been proposed to asses parameter estimability from the sensitivity matrix \cite{Mclean2012}, the simplest of which is to simply plot the sensitivity co-effecients as a function of time, and visually check for obvious linear relations between the curves.

@@

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3]{Sensitivity_unscaled}
        \caption{Unscaled}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3]{Sensitivity_scaled}
        \caption{Scaled}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Sensitivity co-efficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset. \ref{SensitivityMatrix_unscaled} shows them unscaled, \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$.} 
\label{SensitivityMatrix}   
\end{figure}

@@

We suggest that part of the issue is the step function forcing used in model predictions, and the timescales on which the system responds to it. We hypothesize that the system may have two timescales in it - a fast timescale in which eqs. \ref{eq:s} - \ref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in eqs. \ref{eq:p}, \ref{eq:g}, in which measured fluorescence changes in response to the forcing. If this is the case, then it may be that our experimental data can only probe the system via the fixed point of eqs. \ref{eq:s} - \ref{eq:c} , and that the parameters contained within those equations can only work to set the particular fixed point values of $p$ and $g$ that the system tends toward. If this were the case, the particular shape of the models response curve would be primarily determined by the parameters appearing in eqs. \ref{eq:p}, \ref{eq:g}, \ref{eq:z} - namely $\mu$, $\beta$, $f_{s}$ and $\Theta$. If all the information contained in the parameters in eqs.\ref{eq:s} - \ref{eq:c} can only be expressed in the form of the models fixed point, it would not be surprising if there were large spaces of parameters, all giving the same model fixed point, which were indistinguishable.

>>>>

\begin{figure*}[t]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a `double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus the presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}

@@

In this system, transcription of the designed sRNA and mRNA are placed under the control of promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$ \cite{Lutz1997}. These are in turn controlled by two transcriptional repressors, TetR and LacI, which are naturally present in the strain of \textit{E. coli} considered. These repressors disable the promoter regions, and so by default transcription of both RNA's is turned off, and no protein is produced. 
These repressors can themselves be disabled by the presence of two chemicals, aTc and IPTG, which can be introduced externally into the cell (Fig. \ref{ANDGate}).  So transcription of the two RNA's is indirectly controlled by the presence of two chemicals - if neither is present, sRNA and mRNA transcription is repressed, and no protein is produced. If only one is present, the AND gate remains off, either because there is no mRNA to be translated into protein, or because the mRNA is self repressed. But when both are present, the conformational change discussed above occurs, and protein is produced.

@@

mRNA concentrations in this system shown in Fig. \ref{ANDGate} can be indirectly observed by designing the mRNA to code for GFP \footnote{ The methodology of \cite{Rodrigo2012} only optimises the 5' UTR of the mRNA, so the actual protein being coded for is unimportant.}. Recent experiments have used timelapse microscopy to observe the fluoresence of bacteria which contain the above sRNA-mRNA pair, as they are periodically forced with a varying aTc or IPTG concentration \cite{Jaramillo}.

A single layer of the bacteria are grown in rows of chambers. A medium constantly flows through these chambers, allowing normal feeding of the bacteria, and the introduction of aTc or IPTG. The chambers are monitored with software which traces the position of each cell over time, allowing time series of individual cell fluorescences to be recorded.

\begin{figure}[h]
\centering
\includegraphics[trim = 280 0 0  85 ,page=17 ,clip = true, scale=0.4]{WCPM_Warwick-talk45-220115.pdf}
\centering
\includegraphics[trim = 0 0 0  90 ,page=18 ,clip = true, scale=0.35]{WCPM_Warwick-talk45-220115.pdf}
\caption{A schematic of an experimental setup which allows single cell fluorescences to be recorded over time. Bacteria are grown in chambers, with a constant supply of medium flowing through them, which allows the introduction of aTc and IPTG. Software then traces these cells, allowing fluorescence timesseries to be built.  Image reproduced from \cite{Jaramillo}.}
\label{}
\end{figure}

The data we will consider consists of two sets of individual cell time series, labelled \textit{13\_9} and \textit{14\_7}. They correspond to different experimental runs of the above apparatus, for which IPTG concentration was held constant, at a value assumed large enough to saturate the cell's response, and aTc concentrations were varied periodically. Appendix \ref{Initial  Experimental Data} shows the full datasets, with their forcing functions.

@@

In eqs. \ref{eq:s} - \ref{eq:z}, we propose a simple model consisting of a set of ODE's with mass action kinetics to describe the system \cite{UriAlon}. Its state is given by the vector $(s,m,s:m,c,p,g,z)$, with all other variables representing model parameters. Tables \ref{ModelParameters} and \ref{StateVariables} give complete descriptions of the parameters and state variables.

\begin{align}
\frac{ds}{dt} &= \frac{N\alpha_{T}}{f_{T}} y(t)-(\mu + \delta_{s})s -k_{\mathrm{on}}sm +k_{\mathrm{off}}s:m \label{eq:s}\\
\frac{dm}{dt} &=  \frac{N\alpha_{L}}{f_{L}}x(t)-(\mu + \delta_{m})m -k_{\mathrm{on}}sm +k_{\mathrm{off}}s:m  \label{eq:m}\\
\frac{ds:m}{dt} & = k_{\mathrm{on}}sm  - (k_{\mathrm{off}}+ k_{\mathrm{hyb}})s:m  -(\mu + \delta_{sm} )s:m \label{eq:sm}\\
\frac{dc}{dt} & = k_{\mathrm{hyb}}s:m  -(\mu + \delta_{c})c  \label{eq:c} \\
\frac{dp}{dt} & = \beta m +f_{s}\beta c -(\gamma + \mu + \delta_{g})p - \frac{v_{z}p}{K_{z}+p+g}  \label{eq:p} \\
\frac{dg}{dt} & = \gamma p - (\mu + \delta_{g})g - \frac{v_{z}g}{K_{z}+p+g} \label{eq:g} \\
z &= z_{0} +\frac{g}{\Theta} \label{eq:z}
\end{align}
  Based on the reaction mechanism in Fig. \ref{reactionscheme}, the hybridization of the sRNA and mRNA first into an unstable complex, then a stable one, is modelled in eqs. \ref{eq:s}-\ref{eq:c}. The initial binding is modelled as a reversible reaction with forward and backward rates $k_\mathrm{on}$ and $k_\mathrm{off}$.

\begin{align*}
\mathrm{
sRNA \chemsign+ mRNA\chemrel[$k_\mathrm{on}$][$k_\mathrm{off}$]{<>} sRNA:mRNA\textsubscript{unstable}
}
\end{align*}
after which the stabilization is modelled as an irreversible reaction with rate $k_\mathrm{hyb}$:

\begin{align*}
\mathrm{
sRNA:mRNA\textsubscript{unstable} \chemrel[$k_\mathrm{hyb}$]{->} sRNA:mRNA\textsubscript{stable} 
}
\end{align*}

 In addition, these complexes are given degradation rates, $\delta_{s}$, $\delta_{m}$, $\delta_{sm}$, $\delta_{c}$, and dilutions of chemical concentrations due to cell growth are modelled with a dilution rate $\mu$.

@@

\begin{table}[h]
\renewcommand{\arraystretch}{1.3}
\caption{Model Parameters (those to be estimated shown in bold)}
\label{ModelParameters}
\centering
\begin{tabular}{| l | l | p{0.6\linewidth} |}
\hline \textbf{Parameter} &  \textbf{Units} & \textbf{Definition}  \\
\hline \hline N & & Number of copies of promoter existing on plasmid DNA  \\
\hline $z_{0}$ &  Arbitrary (AU) & Baseline experimental fluorescence  \\
\hline $\alpha_{L}$ & nM/min & Maximal transcription rate of $P_\mathrm{LlacO-1}$ promoter\\
\hline $\alpha_{T}$  &  nM/min  & Maximal transcription rate of $P_{\mathrm{LtetO-1}}$ promoter \\
\hline $f_{L}$ &  & Unitless ratio between repressed and unrepressed $P_\mathrm{LlacO-1}$ transcription rate   \\ 
\hline $f_{T}$ &  & Unitless ratio between repressed and unrepressed $P_{\mathrm{LtetO-1}}$ transcription rate  \\
\hline $\delta_{g}$  & /min  & GFP degradation rate  \\
\hline $\gamma$ &  /min & GFP maturation rate  \\
\hline $v_{z}$ & nM/min & Degradation constant of clpx  \\
\hline $K_{z}$   &   nM/min & Dissociation constant of clpx  \\
\hline $\boldsymbol{\Theta}$  &   nM/AFU & Ratio between GFP concentration and observed fluoresence  \\
\hline $\boldsymbol{\mu}$ &  /min & Dilution rate  \\
\hline $\boldsymbol{\delta_{m}}$ &  /min & mRNA degredation rate  \\
\hline $\boldsymbol{\delta_{s}}$ &  /min & sRNA degredation rate  \\
\hline $\delta_{sm}$ &  /min & Unstable sRNA:mRNA degradation rate  \\
\hline $\delta_{c}$ &  /min & Stable sRNA:mRNA degradation rate  \\
\hline $\boldsymbol{k_{on}}$ &   /min & sRNA:mRNA binding rate \\
\hline $\boldsymbol{k_{off}}$ &  /min & sRNA:mRNA unbinding rate \\
\hline $\boldsymbol{k_{hyb}}$ &  /min & sRNA:mRNA hybridization rate \\
\hline $\boldsymbol{\beta}$ &   /min & Baseline translation rate of repressed mRNA \\
\hline $\boldsymbol{f_{s}}$ & & Ratio of repressed mRNA to unrepressed complex translation rate. \\
\hline
\end{tabular}
\end{table}

@@

Our next goal is to estimate the unknown parameters of this model, given the available fluorescence time series data, by fitting the predicted time series from the model to the data. Typically, this is done by minimising the least squares error between model prediction and the experimental data \cite{Brewer2008,Algorithms2003, Hu2015}. Suppose we have some ODE model of our system

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t)
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. An error between the model prediction and an experimental time series is defined as
\begin{align}
J(\boldsymbol{\theta}) = \sum_{i =1}^{N} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots N$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\min_{\boldsymbol{\theta}} \sum_{i =1}^{N} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function is multimodal, and may be very rugged. \footnote{this is true even if ODE model is linear in its parameters, as is almost the case for us - though the ODE model is linear, the resulting solutions are in general not. A counterexample is the harmonic oscillator.} A local optimisation algorithm will often get stuck in local minima. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular recommends several Evolutionary Algorithms, of which we choose one, the CMA-ES \cite{Hansen2006,Hansen2011}. \footnote{CMA-ES stands for covariance matrix evolutionary strategy. This algorithm generates a population of test points according to a multivariate gaussian distribution. It then ranks the points based on their fitness function scores, and  considers some number of the highest ranked points as a sample from a new multivariate gaussian. Using this sample, it then estimates a new mean and covariance matrix for this gaussian and the process is iterated.}

In order to reduce the dimensionality of our search space, we can perform a literature search for existing values of some of our parameters, simplify our model to remove others, and place bounds on those that remain. Appendix \ref{Parameter literature review} contains a list of parameter values found in the literature, where available, and their reference, as well as initial bounds placed on parameters not found in the literature. To further reduce the search space, we assume that $\delta_m$, $\delta_{sm}$ and $\delta_c$ all take similar values, and set them equal. After this is done, we are left with a 9 dimensional search space, bounded by a hypercube (parameters to be estimated are shown in bold in table \ref{ModelParameters}).

\begin{figure*}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3, clip = true, trim = 80 0 74 0]{13_9_hist.eps}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset.}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.38,clip = true, trim = 60 12 74 0]{14_7_hist.eps}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
    \end{subfigure}

        \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.25, clip = true, trim = 100 0 90 0]{13_9_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{13\_9} dataset. }
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.25, clip = true, trim = 80 0 70 0]{14_7_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{14\_7} dataset.}
    \end{subfigure}

    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.37]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset.}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.37]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
    \end{subfigure}
    \caption{Parameter estimates, inter-parameter correlation values, and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off, and a level which saturates the cell's response \todo[inline]{colour forcings differntly, the 14 7 has the wrong forcing plotted, put a colorbar on the maps.}}
\label{InitialResults}
\end{figure*}

@@

Though the predictions are reasonable, they are substantially worse than the predictions for the data they were trained on. The first reason for this may be that parameter values will vary between experimental runs \cite{Hu2015}. The second may be that the model is overfitting.
\todo[inline]{follow this up properly.}

@@

These problems may stem from structural inadequacies in the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions \cite{Jimenez-Hornero2008,Grewal1976 }. If this is the case, no amount of experimental data will allow us to estimate parameters, and we must consider reformulating the model. \footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product $\beta_{1}\beta_{2}$ can ever be estimated, a problem no amount of data can fix.}

Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in general this is not true. \footnote{\cite{Beck} gives an example in which a parameter only affects model predictions after several hours, though others will affect it at all timescales - in this example, if we only took data for a few minutes, the parameter would be inestimable, but in principle it is not.} In this case, parameter estimates may be improved by taking data in more varied experimental conditions, and attempting to observe as many components of the model output as possible.

@@

It can be shown that if sensitivity coefficients are linearly dependent over the range of observation values, the associated parameters cannot be simultaneously estimated \cite{Beck}. Related to this fact, a number of measures have been proposed to assess parameter estimability from the sensitivity matrix \cite{Mclean2012}, the simplest of which is to plot the sensitivity coefficients as a function of time, and visually check for obvious linear relations between the curves.

@@

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3]{Sensitivity_unscaled}
        \caption{Unscaled}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3]{Sensitivity_scaled}
        \caption{Scaled}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Sensitivity coefficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset. \ref{SensitivityMatrix_unscaled} shows them unscaled, \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$.} 
\label{SensitivityMatrix}   
\end{figure}

@@

We suggest that part of the issue is the step function forcing used in model predictions, and the timescales on which the system responds to it. We hypothesize that the system may have two timescales in it - a fast timescale in which eqs. \ref{eq:s} - \ref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in eqs. \ref{eq:p}, \ref{eq:g}, in which measured fluorescence changes in response to the forcing. If this is the case, then it may be that our experimental data can only probe the system via the fixed point of eqs. \ref{eq:s} - \ref{eq:c} , and that the parameters contained within those equations can only work to set the particular fixed point values of $p$ and $g$ that the system tends toward. If this were the case, the particular shape of the models response curve would be primarily determined by the parameters appearing in eqs. \ref{eq:p} - \ref{eq:z} - namely $\mu$, $\beta$, $f_{s}$ and $\Theta$. If all the information contained in the parameters in eqs.\ref{eq:s} - \ref{eq:c} can only be expressed in the form of the model's fixed point, it would not be surprising if there were large spaces of parameters, all giving the same model fixed point, which were indistinguishable.
